\section{Related work}
\label{sec:related}

 
The early work \cite{yuTaxonomy} proposes a generic architecture for grid workflow systems that is based on the workflow reference model defined by the Workflow Management Coalition \cite{wfRefModel}. 
Deelman et al. \cite{deelmanFGCS} have further characterized the features of workflow systems and proposed a general architecture?? 
These two works, however, do not mention science gateways.

Shahand et al. \cite{shahand:2015ab} presents the Science Gateway Canvas, which is a business reference model where the most relevant functional components are organized into functional groups. Although workflow management is mentioned as a possible  component to coordinate distributed computations in science gateways, the paper does not comment on architecture to achieve this.

Olabarriaga et al. \cite{olabarriaga2014} presents a user-centered view of the ecosystem of science gateways with focus on workflow management. Their proposed layered system architecture includes gateway, workflow engine and distributed infrastructure components, similarly to the "service invocation" pattern. However other patterns are not identified.





@inproceedings{shahand:2015ab,
	Address = {Portland, OR, USA},
	Author = {Shahand, Shayan and van Kampen, Antoine H. C. and Olabarriaga, S\'{\i}lvia D.},
	Booktitle = {{Proceedings of the Science of Cyberinfrastructure: Research, Experience, Applications and Models}},
	Doi = {10.1145/2753524.2753527},
	Series = {SCREAM'15},
	Title = {{Science Gateway Canvas: A business reference model for Science Gateways}},
	Year = {2015},
	Bdsk-Url-1 = {http://dx.doi.org/10.1145/2753524.2753527}}

@inproceedings{olabarriaga2014,
	Author = {Olabarriaga, Silvia and Pierantoni, Gabrielle and Taffoni, Giuliano and Sciacca, Eva and Jaghoori, Mahdi and Korkhov, Vladimir and Castelli, Giuliano and Vuerli, Claudio and Becciani, Ugo and Carley, Eoin and others},
	Booktitle = {e-Science (e-Science), 2014 IEEE 10th International Conference on},
	Organization = {IEEE},
	Pages = {298--305},
	Title = {Scientific Workflow Management--For Whom?},
	Year = {2014}}